---
title: "Importing HILDA Files v01"
author: "Craig Whiffin"
date: "17/09/2019"
output:
  html_document: default
---

```{r setup, include=FALSE}
# The knitr optins determine how eac chunk of code is handled in an RMarkdown file
# See https://yihui.name/knitr/options/
knitr::opts_chunk$set(echo = TRUE)

# Reset the environment to avoid using old dataframes, variables etc.
rm(list=ls())

# This section is reserved for adding libraries used throughout the file. Any new libraries will be added here
# at the beginning of the code so they are easy to find / amend in the future.
library(tidyverse)        # loads lots of useful packages
library(haven)            # converts the HILDA files from SPSS format (system data file format) - also reads others
                          # proprietary format for the SPSS statistical software application

# Locating the data files to import - allocate the starting point

# Path for HILDA files using secure PC at home
paths <- c(
  "E:/IN27/HILDA/2. SPSS 170c (Zip File 1 of 2 - Combined Data Files)",
  "E:/IN27/HILDA/2. SPSS 170c (Zip File 2 of 2 - Other Data Files)"
  )

# Path for HILDA files using secure laptop
# TBC when next using laptop


```
# Background
The [Household, Income and Labour Dynamics in Australia (HILDA) Survey, GENERAL RELEASE 17 (Waves 1-17)] (https://dataverse.ada.edu.au/dataset.xhtml?persistentId=doi:10.26193/PTKLYP) were provided by the Australian Data Archive (ADA). This document searches for the data files from a specified directory and converts each of them to an RData file for convenient and better performance. The specified paths are:

```{r paths, comment = NA, echo = FALSE}
paths
```

The RData files will be found in the `..\HILDA RData` directory.

## Find and list the HILDA data files to be converted
The files share a '.sav' extension, so only these need to be located and converted.

```{r find and store the location of the files}
# Set up some variables for lists of file and path names
pathname_list <- list()
filename_list <- list()

for(p in paths){
  #get the entire filename including path
  pathname_list[[p]] <- list.files(path = p, pattern="*.sav", full.names = TRUE)
  #get only the filename of the specific file
  filename_list[[p]] <- list.files(path = p, pattern="*.sav", full.names = FALSE)
}

# Now we have the file and path names, let's put them in a table where we can access them at leisure
HILDA_files <- tibble(
  filename=unlist(filename_list, use.names = FALSE),
  pathname=unlist(pathname_list, use.names = FALSE)
)

# Remove some variables no longer needed
rm(p, paths, pathname_list, filename_list)

```

The resulting HILDA_files table looks like this:
```{r show HILDA_files table, echo=FALSE, comment = FALSE}
HILDA_files
```

# Convert files to RData file format (`.rda`)
### Set up the names and destination for new .rda files
```{r set up rda file destination}
# We can use the HILDA_files table to open each file and convert to RData in an specified output directory
HILDA_files %>%
  mutate(
    basename = str_remove(filename, ".sav$"),       # Remove the .sav extension from each file
    Robjname = make.names(basename),                # Make syntactically valid names out of character vectors.
    Rfilename = str_c(Robjname, ".rda"),            # Append .rda to the new valid name
    Rpathname = str_c("E:/IN27/HILDA/data/", Rfilename)        # This is where we'll put the new rdata (.rda) files
  ) -> HILDA_files
```

### Open each file and convert to the Rdata file format ('.rda`)
Using the haven package, we'll read in each file and perform the conversion. The HILDA_files table will be used
to inform us of the source path and files and the destination path and files.

```{r open and convert each file}
start_time_overall <- Sys.time()
for (i in 1:nrow(HILDA_files)){
  start_time <- Sys.time()         # let's time how long it takes - start the clock here
  cat("Importing ", HILDA_files$filename[i], "...")        # some information for the user
  
  tmp <- read_sav(HILDA_files$pathname[i], user_na=TRUE)      # Import SPSS file as tmp file using
                                                              # user defined NA codes
  assign(HILDA_files$Robjname[i], tmp)           # Rename tmp to the new robjname file
                                                 # Assign() - Assign a value to a name in an environment.
  
  cat("Writing ", HILDA_files$Rfilename[i], "...")   # More information for the user
  save(
    list=c(HILDA_files$Robjname[i]),
    file=HILDA_files$Rpathname[i]
    )
    
  rm(list=c(HILDA_files$Robjname[i]), tmp)     # Tidy up memory
  cat("Completed.")                            # Confirm completed
  print(Sys.time() - start_time)               # Show how long it took
}
  
print(Sys.time() - start_time_overall)         # Show how long it took overall
rm(start_time, start_time_overall, i)
```


